---
title: 'Socially Vulnerable Populations and Susceptibility to Damages'
subtitle: 'Spatial Autocorrelation'
geometry: margin=1in
output:
  github_document: default
  pdf_document: default
  toc: yes
  number_sections: yes
fontsize: 11pt
knit: (function(inputFile, encoding) {
  rmarkdown::render(
    inputFile, encoding = encoding,
    output_format = c('github_document', 'pdf_document'))
    })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(rgdal)      # for readOGR and others
library(sp)         # for spatial objects
library(spdep)      # for modeling
library(dplyr)      # for working with data frames
library(data.table) # for working with data.tables
library(ggplot2)    # for plotting 
library(maptools)   # for plotting 
library(tmap)       # for plotting
library(tigris)     # for geo_join
library(rgeos)      # for spatial objects
library(spatialreg) # for spatial regression
library(stargazer)  # for regression models
```

### Load the shapefiles as SpatialPolygonsDataFrame

```{r}
# Read spatial data
pr_tracts <- readOGR(dsn="../data/census-tract/shapefiles", layer = "cb_2017_72_tract_500k")

# Convert the GEOID to a character
pr_tracts@data$GEOID<-as.character(pr_tracts@data$GEOID)
```

### Load FEMA data

```{r}
d_fema <- fread(file = '../data/open-fema/FEMA-Large-Tract-Demographics-WindSpeed-PR.csv', encoding = 'UTF-8')

# Convert variables names to lower case
names(d_fema) <- tolower(names(d_fema))

d_fema <- select(d_fema, censustractid, county, below_poverty_rate, bachelors_degree_rate, unemployed_labor_rate, owner_occupied_rate, built_1979_or_earlier_rate, waterlevel, pwg_mph, pwg_saffir_simpson, damageamount_cap)
d_fema$censustractid <- as.character(d_fema$censustractid)

head(d_fema)
```

### Join the SpatialPolygonsDataFrame with the Dema dataframe

```{r}
pr_tracts_demo <- geo_join(pr_tracts, d_fema, by_sp="GEOID", by_df="censustractid", how="inner")
nrow(pr_tracts_demo)
```

### Create neighbors

```{r}
nbq <- poly2nb(pr_tracts_demo)
nblist <- nb2listw(nbq, zero.policy = T, style="W")
summary(nbq)
```

### Plot using the plot function

```{r}
plot(pr_tracts_demo, border="grey60")
plot(nbq, coordinates(pr_tracts_demo), add=TRUE, pch=".")
```

### Check Moran's I for below_poverty_rate

```{r}
moran.test(pr_tracts_demo$below_poverty_rate, nblist, zero.policy = T, randomisation = FALSE, na.action = na.omit)
```

### Check Moran's I for owner_occupied_rate

```{r}
moran.test(pr_tracts_demo$owner_occupied_rate, nblist, zero.policy = T, randomisation = FALSE, na.action = na.omit)
```

### Check Moran's I for pwg_saffir_simpson

```{r}
moran.test(pr_tracts_demo$pwg_saffir_simpson, nblist, zero.policy = T, randomisation = FALSE, na.action = na.omit)
```

## Linear Regression

```{r}
# OLS regression
lm_model <- lm(formula=log1p(damageamount_cap) ~ pwg_saffir_simpson + log1p(waterlevel) + below_poverty_rate + bachelors_degree_rate + unemployed_labor_rate + built_1979_or_earlier_rate + owner_occupied_rate, data = pr_tracts_demo)
lm_model$AIC <- AIC(lm_model)
summary(lm_model)
shapiro.test(lm_model$residuals)
```

## Spatial Dependencies

```{r}
# Moran test for spatial correlation in residuals
(moran_model <- lm.morantest(lm_model, listw = nblist, zero.policy = T, alternative = 'two.sided'))
```

## Removing NAs for Plots

```{r}
# Census tracts with NAs
d_fema[rowSums(is.na(d_fema)) > 0, ]

# Remove NAs from spatial
pr_tracts_demo_clean <- pr_tracts_demo[!(pr_tracts_demo@data$GEOID %in% c('72037160100', '72113073008')), ]
```

```{r}
# OLS - NA removed
lm_model_clean <- lm(formula=log1p(damageamount_cap) ~ pwg_saffir_simpson + log1p(waterlevel) + below_poverty_rate + bachelors_degree_rate + unemployed_labor_rate + built_1979_or_earlier_rate + owner_occupied_rate, data = pr_tracts_demo)

# Add residuals
pr_tracts_demo_clean@data$residuals <- residuals(lm_model_clean)
```


## Plotting OLS Residuals

```{r, fig.width=10, fig.height=3.5}
# Residual plot
spplot(pr_tracts_demo_clean, c('residuals'), colorkey=list(space='bottom'),
       main = 'OLS Residuals')
```

```{r, warning=FALSE}
# Adjust bounding box
bbox_new <- st_bbox(pr_tracts_demo_clean) # current bounding box

# Range of values
xrange <- bbox_new$xmax - bbox_new$xmin # range of x values
yrange <- bbox_new$ymax - bbox_new$ymin # range of y values

# Options for box adjustments
bbox_new[1] <- bbox_new[1] + (0.21 * xrange) # xmin - left
bbox_new[3] <- bbox_new[3] - (0.01 * xrange) # xmax - right
# bbox_new[2] <- bbox_new[2] - (0.25 * yrange) # ymin - bottom
# bbox_new[4] <- bbox_new[4] + (0.2 * yrange) # ymax - top

# Make bounding box a sf polygon
bbox_new <- bbox_new %>%
  st_as_sfc()
```

```{r, fig.width=10, fig.height=3.5, warning=FALSE}
# Plot residuals
tm_shape(pr_tracts_demo_clean, bbox = bbox_new) + 
  tm_fill('residuals', title = 'Residuals', palette = '-RdBu', midpoint = 0) +
  tm_borders(alpha = 0.1) +
  tm_layout(main.title = 'OLS Residuals', main.title.size = 1,
            legend.position = c('left', 'bottom'), legend.title.size = 0.9)
```

## Spatial LM Tests

```{r}
# Run LaGrange multiplier tests
lmtests <- lm.LMtests(lm_model, nblist, zero.policy = T, test = c('LMerr','LMlag','RLMerr','RLMlag','SARMA'))
summary(lmtests)

# Both LMerr and LMlag are significant and so must check robust versions
# RLMerr is not significant (p=0.082), indicating spatial error should be used
```

## Spatial Regression

```{r}
# Spatial regression - error model
sem_model <- errorsarlm(formula = log1p(damageamount_cap) ~ pwg_saffir_simpson + log1p(waterlevel) + below_poverty_rate + bachelors_degree_rate + unemployed_labor_rate + built_1979_or_earlier_rate + owner_occupied_rate, data = pr_tracts_demo, listw = nblist, zero.policy = T)

summary(sem_model)

Hausman.test(sem_model)
```

```{r, results='asis', echo=FALSE, warning=FALSE}
# Regression tablefont.size = 'small'
stargazer(lm_model, sem_model, 
          title = 'Social Vulnerability and Susceptibility to Damages',
          header = FALSE, type = 'text', model.names = FALSE, 
          dep.var.labels = 'Damage Amount per Capita (Log)',
          covariate.labels = c('Peak Wind Gusts', 'Water Level (log)','Below Poverty Rate', 'Bachelor Degree Rate',
                               'Unemployed Labor Rate', 'Homes built 1979 or Earlier Rate', 'Owner Occupied Rate'),
          keep.stat=c("aic", "n"),
          column.labels = c('OLS', 'Spatial Error'))
```

